---
title: "Group 2"
author:
  - Silver Lee-A-Fong
  - Jakub Lewkowicz
  - Domantas Pukelevičius
date: "2023-03-02"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 1. Trees
#### a)
In treeVolume data set we can distinguish response variable "Volume" and explanatory variables such as "type", "height" and "diameter". We performed ANOVA with response variable "Volume" and only one explanatory variable "type" to analyse if type of a tree impacts its volume. Based on p-value from ANOVA table, we can observe that tree volume is not significantly impacted by a tree type.
```{r}
df <- read.csv("Data/treeVolume.txt",header = TRUE, sep = "")
df$type = as.factor(df$type)
anova_m = lm(volume ~ type, data=df);anova(anova_m)
```
We split data into tree categories and compared means of those two sample groups by performing t-test. Such t-test can be related to ANOVA performed above, because it also compares significant difference between volumes in those two groups. Summarizing t-test p-value implies that volume is not impacted by type of a tree.

```{r}
beech_samples = df[df$type == "beech", "volume"]
oak_samples = df[df$type == "oak", "volume"]
t.test(beech_samples, oak_samples)
```

Estimated volumes of those two tree types can be expressed as sample mean for a particular group and standard error.
\[\bar{X} = \frac{1}{n}\sum_{i=1}^{n}X_i\]
\[SE = \frac{\sigma}{\sqrt{n}}\]
```{r}
beech_mean = mean(beech_samples); oak_mean = mean(oak_samples)
se_beech = sd(beech_samples) / sqrt(length(beech_samples))
se_oak = sd(oak_samples) / sqrt(length(oak_samples))
beech_mean; se_beech
oak_mean; se_oak
```

#### b)

We performed ANOVA where we included both height and diameter as explanatory variables and volume as response variable. From ANOVA table we can observe that both factors have significant impact on tree's volume.

```{r}
anova_m = lm(volume ~ height + diameter, data=df); anova(anova_m)
```
We also performed different ANOVAs on groups separated by types. In this test we wanted to test influence of diameter and height separately on those two groups. From ANOVA tables where our explanatory variable was diameter, we can observe that volume is influenced in non significantly different way for both groups. Contrary, from ANOVA tables where height was explanatory variable, we can observe that it does not influence oak's volume in a significant way. Beech type tree's height influences it's volume in a significant way.
```{r}
anova_beech_diameter = lm(beech_samples ~ df[df$type == "beech", "diameter"]); anova(anova_beech_diameter)
anova_oak_diameter = lm(oak_samples ~ df[df$type == "oak", "diameter"]); anova(anova_oak_diameter)

anova_beech_height = lm(beech_samples ~ df[df$type == "beech", "height"]); anova(anova_beech_height)
anova_oak_height = lm(oak_samples ~ df[df$type == "oak", "height"]); anova(anova_oak_height)
```

#### c)

Based on previously achieved results we can draw following conclusions:
  - Type does not impact tree's volume.
  - If we take all samples into consideration both height and diameter impact tree's volume significantly.
  - If we analyse height and diameter influence on tree's volume grouped by tree's type, we can observe that height does not influence oak's volume in a significant way.
  
We can observe that height and diameter are correlated for beech type trees, while they are not for oak type trees. This observation can also explain why height does not impact oak's volume in a significant way (what we observed previously).
```{r}
par(mfrow=c(1, 2))
plot(df[df$type == "beech", "diameter"], df[df$type == "beech", "height"])
plot(df[df$type == "oak", "diameter"], df[df$type == "oak", "height"])
cor.test(df[df$type == "beech", "diameter"], df[df$type == "beech", "height"])
cor.test(df[df$type == "oak", "diameter"], df[df$type == "oak", "height"])
```
Using the resulting model we predicted volume of a tree, based on the avaregae diameter.
```{r}
avg_diameter = mean(df$diameter)
model = lm(volume ~ diameter, data=df);
predict(model, data.frame(diameter = c(avg_diameter)))
```

#### d)
We decided to compare two different linear models. In both volume is a response variable, both have height and diameter as explanatory variable. Second model additionaly includes interaction between diameter and height, because those factors might be dependent on each other, therefore we decided to test it. P-value from ANOVA table confirms that adding interaction (between diameter and height) to a model2, significantly improves its fit over model1.
```{r}
model1 = lm(volume ~ diameter + height, data=df)
model2 = lm(volume ~ diameter + height + diameter*height, data=df)
anova(model1, model2)
```

### Excercise 3. Titanic

#### a)
From the table we can observe that more women, regardless of age, survived compared to men. Most of the men that survived were young and under the age of 14. We can show this with a barplot, in the first barplot we can observe that most of the survivors were female and the highest amount of survivors falls in the range of 15-35. We also observe that the 1st PClasshas the most people survived, followed by 3rd and 2nd at last. We can see from logistic regression that PCLass2nd, PClass3rd, Age, Sexmale are significatant for the survival status.

```{r}
df <- read.csv("Data/titanic.txt", header=TRUE, sep="")


tot = xtabs(~PClass+Age+Sex, data=df)
tot.c=xtabs(Survived~PClass+Age+Sex, data=df)
round(tot.c/tot, 2)

tot2 = xtabs(Survived~Sex+Age, data=df)
tot_partial = xtabs(~Sex+Age, data=df)
barplot(tot2, tot_partial, xlab="age", ylab="Total Survived", legend=TRUE)

tot_pclass=xtabs(Survived~PClass, data=df)
tot_pclass_partial=xtabs(~PClass, data=df)
barplot(tot_pclass, tot_pclass_partial, xlab="Pclass", ylab="Total Survived")


# Fit a logistic regression model
titanic_model_a = glm(Survived ~ PClass + Age + Sex, data=df, family="binomial")
summary(titanic_model_a)
```

#### b)

For the interaction between Age and PClass, both Age and PClass are significant which means there exists an interaction. For Age and Sex, only Sex is significant. Indicating that Age is insignificant. The best model from these two should be Age with PClass. We can observe that the probability for survival is significant with $0.048$.

```{r}

# Age with PClass
drop1(glm(Survived ~ Age+PClass, data=df), test="Chisq")

# Age with Sex
drop1(glm(Survived ~ Age+Sex, data=df), test="Chisq")


# Age with PClass factored
df$Age = factor(df$Age)
df$PClass = factor(df$PClass)
apclass = glm(Survived ~ Age+PClass, data=df)
summary(apclass)
```


#### c)



#### d)



#### e)



### Excercise 4. Military coups

#### a)

By performing Poisson regression on the full data set (miltcoup as response variable and oligarchy, pollib, parties, pctvote, popn, suze, numelec, numregim as explanatory variables) we can observe (based on p-values) that only some variables are significant: oligarchy, pollib and parties. Remaining variables do not seem to have significant impact on the model.

```{r}
df <- read.csv("Data/coups.txt", header=TRUE, sep=" ")
model_full = glm(miltcoup ~ oligarchy + pollib + parties + pctvote + popn + size + numelec + numregim, data=df, family = "poisson")
summary(model_full)
```

#### b)

Next following step down approach variable "numelec" is being eliminated due to the highest p-value. Variable represents total number of legislative and presidential elections.
```{r}
model = glm(miltcoup ~ oligarchy + pollib + parties + pctvote + popn + size + numregim, data=df, family = "poisson")
summary(model)
```
Next we eliminate "numregim:
```{r}
model = glm(miltcoup ~ oligarchy + pollib + parties + pctvote + popn + size, data=df, family = "poisson")
summary(model)
```
Next we eliminate "size":
```{r}
model = glm(miltcoup ~ oligarchy + pollib + parties + pctvote + popn, data=df, family = "poisson")
summary(model)
```
Next we eliminate "popn":
```{r}
model = glm(miltcoup ~ oligarchy + pollib + parties + pctvote, data=df, family = "poisson")
summary(model)
```
finally we eliminate "pctvote" ending up with the model were are explanatory variables are significant:
```{r}
model = glm(miltcoup ~ oligarchy + pollib + parties, data=df, family = "poisson")
summary(model)
```
We can compare new model with eliminated variables and old model that uses variables by performing likelyhood ratio test.
```{r}
anova(model_full, model, test="LRT")
```
#### c)

By using the model from b) to predict the number of coups for a hypothetical country for all the three levels of political liberalization and averages of remaining characteristics (parties and oligarchy), we can clearly observe that level of political liberalization significantly impacts results. The lower political liberalization factor, the higher predicted number of coups.

```{r}
avg_oligarchy = mean(df$oligarchy)
avg_parties = mean(df$parties)
d0 = data.frame(oligarchy = c(avg_oligarchy), parties=c(avg_parties), pollib=c(0))
d1 = data.frame(oligarchy = c(avg_oligarchy), parties=c(avg_parties), pollib=c(1))
d2 = data.frame(oligarchy = c(avg_oligarchy), parties=c(avg_parties), pollib=c(2))
response = numeric(3)
response[1] = predict(model, newdata = d0, type="response")
response[2] = predict(model, newdata = d1, type="response")
response[3] = predict(model, newdata = d2, type="response")
pollib = c(0, 1, 2)
plot(pollib, response)
```
